---
title: "Datasaurus or Why Visualizations Are Important"
author: "Stephen Kaluzny"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("dplyr")
library("ggplot2")
```
This note uses a group of datasets that were recently released
to show the importance of visualizations when doing a data analysis.
It illustrates how relying on significant bivariate correlations can lead one astray.

The datasets are a larger and more modern version of what is know
as Anscombe's Quartet.
Anscombe's quartet consisted of 4 bivariate datasets with the same mean,
standard deviation and correlation
but with distinctly different scatterplots.

## The Data

The data is now available as an R package, `datasauRus`, on CRAN.

```{r package, message=FALSE}
if(!suppressWarnings(require("datasauRus", quietly=TRUE, character.only=TRUE))) {
  install.packages("datasauRus")
  library("datasauRus")
}
```
There are 13 datasets consisting of x, y, variable pairs.

We will use the combined (stacked), 3 column version of the datasets
(called `datasaurus_dozen`)
here to easily compute statistics
and plots for all the datasets at once using the `tidyverse`.

The columns have roughly the same univariate statistics:
```{r univariate}
datasaurus_dozen %>%
  group_by(dataset) %>%
  summarise(MeanX=mean(x), SdX=sd(x), MeanY=mean(y), SdY=sd(y))
```

The (pearson) correlations between the pairs are also roughly the same (-.06)
and none of the correlations are considered significant
when tested with `cor.test`
```{r correlation}
datasaurus_dozen %>%
  group_by(dataset) %>%
  summarise(Corr=cor(x, y), Pvalue=cor.test(x, y)$p.value)
```

## Visualization

The surprising result are the scatterplots for each dataset.
Clearly, there is some strong structure in each dataset
yet the bivariate correlations gave no indication of this.

```{r plot}
datasaurus_dozen %>%
  ggplot(aes(x=x, y=y, color=dataset)) +
    geom_point(size=0.5) +
    facet_wrap(~ dataset) +
    guides(color=FALSE)
```

## Appendix
The the `datasauRus` package was originally posted to Github.
To install that version in R:
```{r datasaurus_package, eval=FALSE}
devtools::install_github("stephlocke/datasauRus")
```

The package contains information and references to how the datasets
were created.
